Bootstrap learning via modular concept discovery

### 2013-04-22 drafts/ijcai2013/camera_ready/Dechter.IJCAI13.CameraReady.tex
Spellcheck and minor formatting changes.

\maketitle
\begin{abstract}
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer the learner effectively. How can such a learner ever
 get off the ground? A common intuition is that if the solutions to
 these problems share a common structure, and the learner can solve
 some simple problems by brute force, it should be able to extract
 useful components from these solutions and, by composing them,
 explore the solution space more efficiently. Here, we formalize this
 intuition, where the solution space is the space of typed functional
 programs and the gained information is stored as a stochastic grammar
 over programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and Boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is able to solve almost none of the
 posed problems in the procedure's first iteration, it rapidly becomes
 able to solve a large number by gaining abstract knowledge of the
 structure of the solution space.
\end{abstract}

@@

Domain specific knowledge provides this structure. In local search, it
defines the similarity metric between solutions; in heuristic search
it provides the heuristic. Here we ask how to bootstrap domain
specific knowledge and extract structure from the space of possible
solutions by attempting to solve multiple related problems in a given
domain at once. Our approach is motivated by the observation that many
interesting problems -- such as discovering scientific theories,
developing new technologies, learning to play a new instrument -- have
solutions that are natrually specified as programs.

@@

\section{Related Work}
The idea of discovering and using reusable subcomponents is part of an
old tradition in AI. For instance, it is one of the central themes in
Herb Simon's 1969 ``The Sciences of the
Artificial''~\shortcite{simon1996sciences}. Rendell's 1985
paper~\shortcite{DBLP:conf/ijcai/Rendell85}, ``Substantial Constructive
Induction Using Layered Information Compression: Tractable Feature
Formation in Search,'' presented within a classical search paradigm
the idea that compression provides a heuristic for constructing good
representations. Within the Inductive Logic Programming (ILP)
literature, researchers have explored ``Constructive Induction,''
generating new and reusable logic programming
terms~\cite{DBLP:conf/ijcai/Muggleton87}. Later, predicate invention
was also applied in the multitask
setting~\cite{DBLP:conf/ilp/KhanMP98}. We apply these ideas to
learning functional programs and believe that the modularity and
compositionality afforded by this representation is necessary for
these ideas to be successful.

@@

%% %% But suppose that you have very little knowledge of a domain. Your
%% %% intial representation of the domain is poor: ``nearby'' points in the
%% %% solutions space ellicit wildly different outcomes in unpredictable
%% %% ways. Or perhaps nearby points ellicit the same outcome and its not a
%% %% good one, so you don't know which way to go. Many learning problems,
%% %% especially those whose solutions are best described as programs, have
%% %% these qualities: imagine learning to synthesize new sentences

@@

Our goal is to solve the following minimal feedback multi-task program
induction problem: suppose we are given a set of \textbf{tasks}
$T=\{t_k\}_{k=1}^K$ where each task is a function $t_k : \mathcal{L}
\rightarrow \{0, 1\}$ and where $\mathcal{L}$ is the set of
expressions in our language.  We say that expression $ e \in
\mathcal{L}$ \textbf{solves} $t_k$ if $ t_k(e) = 1$. Our goal is to
solve as many of the tasks in $T$ as we can.

We propose the \textbf{E.C. algorithm}, an iterative procedure to solve this
problem. It will be useful in presenting the algorithm to define the
following terminology: a \textbf{frontier} is the finite set of
expressions that the algorithm considers in any one iteration. The
\textbf{frontier size} $N$ is the prespecified number of elements in
the frontier. We will say that a task is \textbf{hit} by the frontier
if there is a expression in the frontier that solves it.

@@

Given a set of chosen solution expressions, which we will call the
\emph{solution set}, want to re-estimate our stochastic grammar. Our
inspiration for doing this is the Neville-Manning algorithm for
grammar-based compression~\cite{nevill1997identifying}. The idea of
that compression algorithm is to compress any string by creating a
grammar for that string such that a) no two symbols appear more than
once in the grammar and b) every rule is used more than once. From a
set of expressions, we generate a grammar according to these criteria.

@@

\setcounter{subfigure}{0} 
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\linewidth]{figures/symreg3frontiers.pdf}
\caption{Learning curves as a function of frontier size.  As frontier
  size is increased, curves plateau closer to $100\%$ performance. A
  baseline search over $150000$ expressions only hits $3\%$ of the
  tasks (dashed pink line). \label{fig:symregLearning}}
\end{figure}
\begin{figure}[ht]
\includegraphics[width=1.07\linewidth]{figures/symRegDiffCurricula.pdf}
\caption{How do different task sets affect learning curves? Learning
  curves at frontier size of $10000$ for different task
  sets.} 
\label{fig:symregCurricula}

>>>>

\maketitle
\begin{abstract}
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer the learner effectively. How can such a learner ever
 get off the ground? A common intuition is that if the solutions to
 these problems share a common structure, and the learner can solve
 some simple problems by brute force, it should be able to extract
 useful components from these solutions and, by composing them,
 explore the solution space more efficiently. Here, we formalize this
 intuition, where the solution space is that of typed functional
 programs and the gained information is stored as a stochastic grammar
 over programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and Boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is able to solve almost none of the
 posed problems in the procedure's first iteration, it rapidly becomes
 able to solve a large number by gaining abstract knowledge of the
 structure of the solution space.
\end{abstract}

@@

Domain specific knowledge provides this structure. In local search, it
defines the similarity metric between solutions; in heuristic search
it provides the heuristic. Here we ask how to bootstrap domain
specific knowledge and extract structure from the space of possible
solutions by attempting to solve multiple related problems in a given
domain at once. Our approach is motivated by the observation that many
interesting problems -- such as discovering scientific theories,
developing new technologies, learning to play a new instrument -- have
solutions that are naturally specified as programs.

@@

\section{Related Work}
The idea of discovering and using reusable subcomponents is part of an
old tradition in AI. For instance, it is one of the central themes in
Herb Simon's ``The Sciences of the
Artificial''~\shortcite{simon1996sciences}. Rendell's ``Substantial
Constructive Induction Using Layered Information Compression:
Tractable Feature Formation in Search'' presented within a classical
search paradigm the idea that compression provides a heuristic for
constructing good
representations~\shortcite{DBLP:conf/ijcai/Rendell85}. Within the
Inductive Logic Programming (ILP) literature, researchers have
explored ``Constructive Induction,'' generating new and reusable logic
programming terms~\cite{DBLP:conf/ijcai/Muggleton87}. Later, predicate
invention was also applied in the multitask
setting~\cite{DBLP:conf/ilp/KhanMP98}. We apply these ideas to
learning functional programs and believe that the modularity and
compositionality afforded by this representation is necessary for
these ideas to be successful.

@@

%% %% But suppose that you have very little knowledge of a domain. Your
%% %% initial representation of the domain is poor: ``nearby'' points in the
%% %% solutions space elicit wildly different outcomes in unpredictable
%% %% ways. Or perhaps nearby points elicit the same outcome and its not a
%% %% good one, so you don't know which way to go. Many learning problems,
%% %% especially those whose solutions are best described as programs, have
%% %% these qualities: imagine learning to synthesize new sentences

@@

Our goal is to solve the following minimal feedback multi-task program
induction problem: suppose we are given a set of \textbf{tasks}
$T=\{t_k\}_{k=1}^K$ where each task is a function \mbox{$t_k : \mathcal{L}
\rightarrow \{0, 1\}$} and where $\mathcal{L}$ is the set of
expressions in our language.  We say that expression $ e \in
\mathcal{L}$ \textbf{solves} $t_k$ if $ t_k(e) = 1$. Our goal is to
solve as many of the tasks in $T$ as we can.

We propose the \textbf{E.C. algorithm}, an iterative procedure to solve this
problem. It will be useful in presenting the algorithm to define the
following terminology: a \textbf{frontier} is the finite set of
expressions that the algorithm considers in any one iteration. The
\textbf{frontier size} $N$ is the pre-specified number of elements in
the frontier. We will say that a task is \textbf{hit} by the frontier
if there is a expression in the frontier that solves it.

@@

Given a set of chosen solution expressions, which we will call the
\emph{solution set}, want to re-estimate our stochastic grammar. Our
inspiration for doing this is the Nevill-Manning algorithm for
grammar-based compression~\cite{nevill1997identifying}. The idea of
that compression algorithm is to compress any string by creating a
grammar for that string such that a) no two symbols appear more than
once in the grammar and b) every rule is used more than once. From a
set of expressions, we generate a grammar according to these criteria.

@@

\setcounter{subfigure}{0} 
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\linewidth]{figures/symreg3frontiers.pdf}
\caption{Learning curves as a function of frontier size.  As frontier
  size is increased, curves plateau closer to $100\%$ performance. A
  baseline search over $150000$ expressions only hits $3\%$ of the
  tasks (dashed pink line). \label{fig:symregLearning}}
\end{figure}
\begin{figure}[ht]
\includegraphics[width=1.07\linewidth]{figures/symRegDiffCurricula.pdf}
\caption{How do different task sets affect learning curves? Learning
  curves at a frontier size of $10000$ for different task
  sets.} 
\label{fig:symregCurricula}
